---
title: "VJ146A2 High p-lit Gate Sensitivity"
subtitle: "Does the VJ/Eskom MLR signal survive looser not-up thresholds?"
author: "Reliable Electrification Planning SA"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: false
    number_sections: false
    fig_caption: true
geometry: margin=0.72in
fontsize: 10pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \usepackage{float}
  - \captionsetup{font=small,labelfont=bf}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7.4,
  fig.height = 4.4,
  dpi = 300,
  dev = "png",
  fig.pos = "H",
  out.width = "100%",
  fig.path = "../figures/vj146a2-high-p-lit-gate-sensitivity/"
)

suppressPackageStartupMessages({
  library(arrow)
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(lmtest)
  library(patchwork)
  library(scales)
  library(sandwich)
  library(sf)
  library(tidyr)
})

sf::sf_use_s2(FALSE)
data.table::setDTthreads(0L)

find_repo_root <- function(start = getwd()) {
  current <- normalizePath(start, mustWork = TRUE)
  repeat {
    if (
      file.exists(file.path(current, ".here")) &&
        dir.exists(file.path(current, "validation"))
    ) {
      return(current)
    }
    parent <- dirname(current)
    if (identical(parent, current)) stop("Could not locate Reliability-Assessment repo root.")
    current <- parent
  }
}

codebase_path <- find_repo_root()
data_path <- file.path(codebase_path, "validation", "eskom_mlr", "data")

panel_dir <- file.path(codebase_path, "Map Data", "settlement_day_outputs_vj146a2")
panel_files <- file.path(
  panel_dir,
  paste0("settlement_day_vj146a2_cov_yearlykeep_2023-", sprintf("%02d", 1:12), ".parquet")
)
yearly_stats_file <- file.path(
  codebase_path,
  "Map Data", "reliability_outputs_vj146a2",
  "yearly_settlement_stats_2023.parquet"
)
sett_gpkg <- file.path(
  codebase_path,
  "Map Data", "Settlements", "GPKG",
  "south_africa_dre_atlas_settlements_full_col.gpkg"
)
coverage_file <- file.path(
  data_path,
  "vj146a2_2023_demand_gate_coverage.csv"
)
flags_file <- file.path(
  data_path,
  "vj146a2_2023_annual_iqr_excess_lit_day_flags.csv"
)

stopifnot(
  all(file.exists(panel_files)),
  file.exists(yearly_stats_file),
  file.exists(sett_gpkg),
  file.exists(coverage_file),
  file.exists(flags_file)
)

thresholds <- c(0.05, 0.20, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 0.95)

color_pop <- "#9B2D30"
color_dem <- "#2D5F8A"
color_fe <- "#6B6B6B"
color_ref <- "#DDAA33"

fmt_num <- function(x, digits = 2) {
  ifelse(is.na(x), "", formatC(x, format = "f", digits = digits))
}

fmt_p <- function(p) {
  ifelse(p < 0.001, formatC(p, format = "e", digits = 2), formatC(p, format = "f", digits = 4))
}

safe_weighted_mean <- function(x, w) {
  idx <- is.finite(x) & is.finite(w) & w > 0
  if (!any(idx)) return(NA_real_)
  sum(x[idx] * w[idx]) / sum(w[idx])
}

fit_nw <- function(d, y, controls = character()) {
  rhs <- c("shed_share_1_2am_primary", controls)
  f <- as.formula(paste(y, "~", paste(rhs, collapse = " + ")))
  sub <- d[is.finite(d[[y]]) & is.finite(d$shed_share_1_2am_primary), ]
  fit <- lm(f, sub)
  vc <- NeweyWest(fit, lag = 7, prewhite = FALSE, adjust = TRUE)
  ct <- coeftest(fit, vcov. = vc)
  term <- "shed_share_1_2am_primary"
  est <- unname(ct[term, "Estimate"])
  se <- unname(ct[term, "Std. Error"])
  p <- unname(ct[term, "Pr(>|t|)"])
  data.frame(
    n = nobs(fit),
    effect_pp_per_10pp = est * 10,
    se_pp_per_10pp = se * 10,
    ci_low_pp = (est - 1.96 * se) * 10,
    ci_high_pp = (est + 1.96 * se) * 10,
    p_value = p,
    pearson_r = if (length(controls) == 0) {
      cor(sub[[y]], sub$shed_share_1_2am_primary, use = "complete.obs")
    } else {
      NA_real_
    },
    r_squared = summary(fit)$r.squared,
    adj_r_squared = summary(fit)$adj.r.squared,
    mean_outcome = mean(sub[[y]], na.rm = TRUE),
    sd_outcome = sd(sub[[y]], na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}
```

```{r reconstruct-outcomes}
sett_static <- sf::st_read(sett_gpkg, quiet = TRUE) |>
  sf::st_drop_geometry() |>
  transmute(
    settlement_id = as.character(settlement_id),
    demand_weight = as.numeric(demand)
  )

yearly_keep <- arrow::read_parquet(yearly_stats_file) |>
  transmute(
    settlement_id = as.character(settlement_id),
    population_yearly = as.numeric(population),
    electrified_best_yearly = as.integer(electrified_best)
  ) |>
  filter(electrified_best_yearly == 1L)

universe <- yearly_keep |>
  left_join(sett_static, by = "settlement_id") |>
  mutate(
    in_all_demand_universe = is.finite(demand_weight) & demand_weight > 0,
    in_validation_universe = in_all_demand_universe &
      is.finite(population_yearly) &
      population_yearly > 100
  ) |>
  select(settlement_id, population_yearly, demand_weight, in_validation_universe)

panel <- data.table::rbindlist(lapply(panel_files, function(f) {
  as.data.table(arrow::read_parquet(
    f,
    col_select = c("settlement_id", "date", "population", "p_lit_sett")
  ))
}), use.names = TRUE)

panel[, settlement_id := as.character(settlement_id)]
panel[, date := as.Date(date)]
panel[, population := as.numeric(population)]
panel[, p_lit_sett := as.numeric(p_lit_sett)]
panel <- merge(panel, as.data.table(universe), by = "settlement_id", all.x = FALSE)
panel <- panel[in_validation_universe == TRUE]
panel[, population := data.table::fcoalesce(population, population_yearly)]
panel <- panel[is.finite(p_lit_sett)]

daily_list <- lapply(thresholds, function(th) {
  suffix <- sprintf("%03d", as.integer(round(th * 100)))
  panel[
    ,
    .(
      popw = safe_weighted_mean(as.numeric(p_lit_sett < th), population),
      demandw = safe_weighted_mean(as.numeric(p_lit_sett < th), demand_weight)
    ),
    by = date
  ][, `:=`(threshold = th, threshold_suffix = suffix)]
})

daily_long <- data.table::rbindlist(daily_list, use.names = TRUE)
daily_wide <- data.table::dcast(daily_long, date ~ threshold_suffix, value.var = c("popw", "demandw"))

coverage <- read.csv(coverage_file, stringsAsFactors = FALSE) |>
  mutate(
    date = as.Date(date),
    shed_share_1_2am_primary = mlr_sum_mwh_1_2am_primary / contracted_demand_sum_mwh_1_2am_primary,
    month_fe = factor(format(date, "%Y-%m")),
    dow_fe = factor(
      weekdays(date),
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
    )
  )

flags <- read.csv(flags_file, stringsAsFactors = FALSE) |>
  mutate(date = as.Date(date))

removed_dates <- flags |>
  filter(scenario == "annual_q3_plus_1p5_iqr", remove) |>
  pull(date)

analysis <- coverage |>
  filter(gate_tier == "pass", !(date %in% removed_dates)) |>
  left_join(as.data.frame(daily_wide), by = "date") |>
  filter(is.finite(shed_share_1_2am_primary))

sanity_checks <- data.frame(
  metric = c("popw_005", "popw_020", "popw_040", "demandw_005", "demandw_020", "demandw_040"),
  max_abs_diff = c(
    max(abs(analysis$popw_005 - analysis$popw_strict_dark_share), na.rm = TRUE),
    max(abs(analysis$popw_020 - analysis$popw_mostly_dark_share), na.rm = TRUE),
    max(abs(analysis$popw_040 - analysis$popw_relaxed_dark_share), na.rm = TRUE),
    max(abs(analysis$demandw_005 - analysis$demandw_strict_dark_share), na.rm = TRUE),
    max(abs(analysis$demandw_020 - analysis$demandw_mostly_dark_share), na.rm = TRUE),
    max(abs(analysis$demandw_040 - analysis$demandw_relaxed_dark_share), na.rm = TRUE)
  )
)

results <- bind_rows(lapply(thresholds, function(th) {
  suffix <- sprintf("%03d", as.integer(round(th * 100)))
  bind_rows(
    fit_nw(analysis, paste0("popw_", suffix)) |>
      mutate(weighting = "Population", threshold = th, model = "Simple OLS + NW"),
    fit_nw(analysis, paste0("demandw_", suffix)) |>
      mutate(weighting = "Demand", threshold = th, model = "Simple OLS + NW"),
    fit_nw(analysis, paste0("popw_", suffix), controls = c("month_fe", "dow_fe")) |>
      mutate(weighting = "Population", threshold = th, model = "Month + DOW FE + NW"),
    fit_nw(analysis, paste0("demandw_", suffix), controls = c("month_fe", "dow_fe")) |>
      mutate(weighting = "Demand", threshold = th, model = "Month + DOW FE + NW")
  )
})) |>
  mutate(
    gate_label = sprintf("%.2f", threshold),
    gate_factor = factor(gate_label, levels = sprintf("%.2f", thresholds))
  )

simple_results <- results |>
  filter(model == "Simple OLS + NW")

fe_results <- results |>
  filter(model == "Month + DOW FE + NW")

preferred_demand_040 <- simple_results |>
  filter(weighting == "Demand", abs(threshold - 0.40) < 1e-9)

settlement_panel <- merge(
  panel,
  as.data.table(analysis |> select(date, shed_share_1_2am_primary)),
  by = "date",
  all.x = FALSE
)

weighted_quantile <- function(x, w, probs) {
  idx <- is.finite(x) & is.finite(w) & w > 0
  x <- x[idx]
  w <- w[idx]
  if (length(x) == 0) return(rep(NA_real_, length(probs)))
  ord <- order(x)
  x <- x[ord]
  w <- w[ord]
  cw <- cumsum(w) / sum(w)
  sapply(probs, function(p) x[which(cw >= p)[1]])
}

settlement_slope_one_threshold <- function(dt, threshold) {
  dt[, y_tmp := as.numeric(p_lit_sett < threshold)]
  out <- dt[
    ,
    {
      ok <- is.finite(y_tmp) & is.finite(shed_share_1_2am_primary)
      y <- y_tmp[ok]
      x <- shed_share_1_2am_primary[ok]
      n <- length(y)
      pop <- population[which(!is.na(population))[1]]
      dem <- demand_weight[which(!is.na(demand_weight))[1]]
      if (n < 50 || !is.finite(stats::var(x)) || stats::var(x) <= 0) {
        .(
          n_obs = n,
          population = pop,
          demand_weight = dem,
          slope_pp = NA_real_,
          p_value = NA_real_,
          constant_outcome = NA
        )
      } else if (length(unique(y)) <= 1) {
        .(
          n_obs = n,
          population = pop,
          demand_weight = dem,
          slope_pp = 0,
          p_value = NA_real_,
          constant_outcome = TRUE
        )
      } else {
        fit <- stats::lm(y ~ x)
        co <- summary(fit)$coefficients
        .(
          n_obs = n,
          population = pop,
          demand_weight = dem,
          slope_pp = unname(co["x", "Estimate"] * 10),
          p_value = unname(co["x", "Pr(>|t|)"]),
          constant_outcome = FALSE
        )
      }
    },
    by = settlement_id
  ]
  out[, threshold := threshold]
  dt[, y_tmp := NULL]
  out
}

settlement_slopes <- rbindlist(
  lapply(thresholds, function(th) settlement_slope_one_threshold(settlement_panel, th)),
  use.names = TRUE
)

settlement_slope_summary <- settlement_slopes[
  is.finite(slope_pp),
  {
    wq <- weighted_quantile(slope_pp, population, c(0.10, 0.25, 0.50, 0.75, 0.90))
    .(
      settlements_estimable = .N,
      population_estimable = sum(population, na.rm = TRUE),
      unweighted_mean_slope_pp = mean(slope_pp, na.rm = TRUE),
      population_weighted_mean_slope_pp = safe_weighted_mean(slope_pp, population),
      median_slope_pp = median(slope_pp, na.rm = TRUE),
      weighted_p10_slope_pp = wq[[1]],
      weighted_p25_slope_pp = wq[[2]],
      weighted_median_slope_pp = wq[[3]],
      weighted_p75_slope_pp = wq[[4]],
      weighted_p90_slope_pp = wq[[5]],
      share_settlements_positive = mean(slope_pp > 0, na.rm = TRUE),
      share_population_positive = safe_weighted_mean(as.numeric(slope_pp > 0), population),
      share_settlements_sig_positive = mean(slope_pp > 0 & p_value < 0.05, na.rm = TRUE),
      share_population_sig_positive = safe_weighted_mean(as.numeric(slope_pp > 0 & p_value < 0.05), population),
      share_settlements_constant = mean(constant_outcome == TRUE, na.rm = TRUE),
      share_population_constant = safe_weighted_mean(as.numeric(constant_outcome == TRUE), population)
    )
  },
  by = threshold
][
  ,
  gate_label := sprintf("%.2f", threshold)
]

selected_slope_gates <- c(0.05, 0.20, 0.40, 0.60, 0.80, 0.95)
slope_plot_data <- settlement_slopes[
  threshold %in% selected_slope_gates & is.finite(slope_pp)
]
slope_trim <- slope_plot_data[
  ,
  .(
    lo = weighted_quantile(slope_pp, population, 0.01),
    hi = weighted_quantile(slope_pp, population, 0.99)
  ),
  by = threshold
]
slope_plot_data <- merge(slope_plot_data, slope_trim, by = "threshold")
slope_plot_data <- slope_plot_data[slope_pp >= lo & slope_pp <= hi]
slope_plot_data[, gate_label := factor(sprintf("%.2f", threshold), levels = sprintf("%.2f", selected_slope_gates))]
slope_mean_lines <- settlement_slope_summary[
  threshold %in% selected_slope_gates,
  .(threshold, gate_label = factor(sprintf("%.2f", threshold), levels = sprintf("%.2f", selected_slope_gates)), population_weighted_mean_slope_pp)
]
```

# Purpose

This report tests whether the VJ/Eskom validation signal weakens when the daily darkness/not-up gate is pushed above the current `p_lit < 0.40` threshold. The main report validates `0.05`, `0.20`, and `0.40`; this sensitivity adds `0.50`, `0.60`, `0.70`, `0.80`, `0.90`, and `0.95`.

The setup follows the favored specification in [[validation/eskom_mlr/reports/2026-06-03-recommended-vj-eskom-validation-specification.Rmd]]:

- standard sample: `gate_tier == "pass"`
- annual-IQR excess-lit removal: `annual_q3_plus_1p5_iqr`
- retained standard sample: `r nrow(analysis)` nights
- regressor: `shed_share_1_2am_primary`
- denominator: `sum(Eskom MLR from 01:00-02:00 SAST) / sum(RSA Contracted Demand from 01:00-02:00 SAST)`
- inference: Newey-West HAC standard errors with 7-day lag

The temporary reconstruction matches the main-report `0.05`, `0.20`, and `0.40` columns to numerical precision; the largest absolute difference is `r format(max(sanity_checks$max_abs_diff), scientific = TRUE, digits = 3)`.

# Main Pattern

```{r fit-degradation-plot, fig.cap="Validation fit degrades as the p-lit gate becomes too loose. R2 is strongest at low-to-moderate thresholds and falls sharply above 0.60, even though coefficient estimates remain positive."}
plot_r2 <- ggplot(simple_results, aes(x = threshold, y = r_squared, color = weighting, group = weighting)) +
  geom_vline(xintercept = 0.40, linetype = "dashed", color = color_ref, linewidth = 0.55) +
  geom_line(linewidth = 0.85) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = thresholds, labels = sprintf("%.2f", thresholds)) +
  scale_y_continuous(labels = number_format(accuracy = 0.01), limits = c(0, NA)) +
  scale_color_manual(values = c(Population = color_pop, Demand = color_dem)) +
  labs(
    title = "R2 falls after the validated not-up range",
    subtitle = "Simple OLS + Newey-West; dashed line marks the current p_lit < 0.40 not-up proxy",
    x = "Daily threshold: p_lit below gate",
    y = "R2",
    color = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

plot_r <- ggplot(simple_results, aes(x = threshold, y = pearson_r, color = weighting, group = weighting)) +
  geom_vline(xintercept = 0.40, linetype = "dashed", color = color_ref, linewidth = 0.55) +
  geom_line(linewidth = 0.85) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = thresholds, labels = sprintf("%.2f", thresholds)) +
  scale_y_continuous(labels = number_format(accuracy = 0.01), limits = c(0, NA)) +
  scale_color_manual(values = c(Population = color_pop, Demand = color_dem)) +
  labs(
    title = "Correlation also weakens at high gates",
    subtitle = "The signal remains positive but becomes less specific to Eskom MLR",
    x = "Daily threshold: p_lit below gate",
    y = "Pearson r",
    color = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

plot_r2 / plot_r
```

The raw coefficient keeps rising because the outcome becomes broader. A `p_lit < 0.95` event mechanically includes all lower-threshold dark or not-up events plus many additional partial-lit states. Fit quality is the more informative diagnostic here: in the demand-weighted bridge, R2 is `r fmt_num(preferred_demand_040$r_squared, 3)` at `0.40`, but only `r fmt_num(simple_results$r_squared[simple_results$weighting == "Demand" & simple_results$threshold == 0.90], 3)` at `0.90` and `r fmt_num(simple_results$r_squared[simple_results$weighting == "Demand" & simple_results$threshold == 0.95], 3)` at `0.95`.

```{r slope-prevalence-plot, fig.cap="Higher gates have larger raw percentage-point slopes because the outcome is more prevalent, but the increasing slope should not be read as a better validation metric."}
plot_effect <- ggplot(simple_results, aes(x = threshold, y = effect_pp_per_10pp, color = weighting, group = weighting)) +
  geom_vline(xintercept = 0.40, linetype = "dashed", color = color_ref, linewidth = 0.55) +
  geom_line(linewidth = 0.85) +
  geom_point(size = 2.2) +
  geom_errorbar(aes(ymin = ci_low_pp, ymax = ci_high_pp), width = 0.015, linewidth = 0.45, alpha = 0.75) +
  scale_x_continuous(breaks = thresholds, labels = sprintf("%.2f", thresholds)) +
  scale_color_manual(values = c(Population = color_pop, Demand = color_dem)) +
  labs(
    title = "Raw slopes rise with broader outcomes",
    subtitle = "Effect is percentage-point change in threshold share for +10 pp Eskom MLR/contracted",
    x = "Daily threshold: p_lit below gate",
    y = "Effect per +10 pp MLR/contracted",
    color = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

plot_mean <- ggplot(simple_results, aes(x = threshold, y = mean_outcome, color = weighting, group = weighting)) +
  geom_vline(xintercept = 0.40, linetype = "dashed", color = color_ref, linewidth = 0.55) +
  geom_line(linewidth = 0.85) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = thresholds, labels = sprintf("%.2f", thresholds)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c(Population = color_pop, Demand = color_dem)) +
  labs(
    title = "High gates are much more prevalent",
    subtitle = "By 0.95, the metric is broad dimness rather than a clean not-up proxy",
    x = "Daily threshold: p_lit below gate",
    y = "Mean daily share",
    color = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

plot_effect / plot_mean
```

```{r fe-comparison-plot, fig.cap="The fixed-effects robustness keeps the same degradation pattern. R2 remains high near 0.40 and falls as the gate approaches 0.95."}
fe_plot_data <- fe_results |>
  mutate(weighting = factor(weighting, levels = c("Population", "Demand")))

ggplot(fe_plot_data, aes(x = threshold, y = r_squared, color = weighting, group = weighting)) +
  geom_vline(xintercept = 0.40, linetype = "dashed", color = color_ref, linewidth = 0.55) +
  geom_line(linewidth = 0.85) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = thresholds, labels = sprintf("%.2f", thresholds)) +
  scale_y_continuous(labels = number_format(accuracy = 0.01), limits = c(0, NA)) +
  scale_color_manual(values = c(Population = color_pop, Demand = color_dem)) +
  labs(
    title = "Month + DOW FE robustness: high gates still degrade",
    subtitle = "The relationship is robust, but very high gates are less selective",
    x = "Daily threshold: p_lit below gate",
    y = "R2",
    color = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

# Results

```{r results-tables}
table_simple <- simple_results |>
  mutate(
    Gate = sprintf("%.2f", threshold),
    `Mean outcome` = percent(mean_outcome, accuracy = 0.01),
    `Effect per +10pp MLR` = paste0(fmt_num(effect_pp_per_10pp, 2), " pp"),
    `NW SE` = paste0(fmt_num(se_pp_per_10pp, 2), " pp"),
    `NW 95% CI` = paste0(fmt_num(ci_low_pp, 2), " to ", fmt_num(ci_high_pp, 2), " pp"),
    `p-value` = fmt_p(p_value),
    `Pearson r` = fmt_num(pearson_r, 3),
    `R2` = fmt_num(r_squared, 3)
  ) |>
  select(Weighting = weighting, Gate, `Mean outcome`, `Effect per +10pp MLR`, `NW SE`, `NW 95% CI`, `p-value`, `Pearson r`, `R2`)

kable(
  table_simple,
  booktabs = TRUE,
  caption = "Simple OLS + Newey-West sensitivity by p-lit gate. The sample and regressor match the favored main-report specification."
)
```

```{r fe-table}
table_fe <- fe_results |>
  mutate(
    Gate = sprintf("%.2f", threshold),
    `Mean outcome` = percent(mean_outcome, accuracy = 0.01),
    `Effect per +10pp MLR` = paste0(fmt_num(effect_pp_per_10pp, 2), " pp"),
    `NW SE` = paste0(fmt_num(se_pp_per_10pp, 2), " pp"),
    `NW 95% CI` = paste0(fmt_num(ci_low_pp, 2), " to ", fmt_num(ci_high_pp, 2), " pp"),
    `p-value` = fmt_p(p_value),
    `R2` = fmt_num(r_squared, 3)
  ) |>
  select(Weighting = weighting, Gate, `Mean outcome`, `Effect per +10pp MLR`, `NW SE`, `NW 95% CI`, `p-value`, `R2`)

kable(
  table_fe,
  booktabs = TRUE,
  caption = "Month + day-of-week fixed-effects sensitivity by p-lit gate."
)
```

# Settlement-Level Heterogeneity

The aggregate threshold sensitivity can remain strong even when many individual settlements have no detectable response. To check whether the settlement-level picture changes as the gate increases, each settlement was separately regressed on the same 01:00-02:00 Eskom MLR share using the standard validation nights. Settlements with at least 50 observations are retained; constant-outcome settlements are assigned zero slopes, matching the convention used in the main report's heterogeneity table.

```{r settlement-slope-summary-table}
settlement_slope_table <- settlement_slope_summary |>
  mutate(
    Gate = sprintf("%.2f", threshold),
    `Estimable settlements` = format(settlements_estimable, big.mark = ","),
    `Weighted mean slope` = paste0(fmt_num(population_weighted_mean_slope_pp, 2), " pp"),
    `Median slope` = paste0(fmt_num(median_slope_pp, 2), " pp"),
    `Weighted median` = paste0(fmt_num(weighted_median_slope_pp, 2), " pp"),
    `Population positive` = percent(share_population_positive, accuracy = 0.1),
    `Population sig. positive` = percent(share_population_sig_positive, accuracy = 0.1),
    `Population constant` = percent(share_population_constant, accuracy = 0.1)
  ) |>
  select(
    Gate,
    `Estimable settlements`,
    `Weighted mean slope`,
    `Median slope`,
    `Weighted median`,
    `Population positive`,
    `Population sig. positive`,
    `Population constant`
  )

kable(
  settlement_slope_table,
  booktabs = TRUE,
  caption = "Per-settlement slope distribution across p-lit gates. Each settlement is separately regressed on Eskom MLR using standard validation nights with at least 50 observations. Constant-outcome settlements are retained with zero slopes."
)
```

```{r settlement-slope-summary-plot, fig.cap="Settlement-level response broadens as the p-lit gate rises. The population-weighted mean slope increases through moderate gates, while the population share with positive slopes rises and the constant-outcome mass falls."}
summary_long <- settlement_slope_summary |>
  transmute(
    threshold,
    `Weighted mean slope` = population_weighted_mean_slope_pp,
    `Population positive` = 100 * share_population_positive,
    `Population significant positive` = 100 * share_population_sig_positive,
    `Population constant outcome` = 100 * share_population_constant
  ) |>
  pivot_longer(-threshold, names_to = "metric", values_to = "value") |>
  mutate(
    panel = ifelse(metric == "Weighted mean slope", "Slope", "Population share"),
    metric = factor(
      metric,
      levels = c("Weighted mean slope", "Population positive", "Population significant positive", "Population constant outcome")
    )
  )

ggplot(summary_long, aes(x = threshold, y = value, color = metric, group = metric)) +
  geom_vline(xintercept = 0.40, linetype = "dashed", color = color_ref, linewidth = 0.55) +
  geom_line(linewidth = 0.85) +
  geom_point(size = 2) +
  facet_wrap(~ panel, scales = "free_y", ncol = 1) +
  scale_x_continuous(breaks = thresholds, labels = sprintf("%.2f", thresholds)) +
  scale_color_manual(values = c(
    "Weighted mean slope" = "#9B2D30",
    "Population positive" = "#2D5F8A",
    "Population significant positive" = "#508C5A",
    "Population constant outcome" = "#7A7A7A"
  )) +
  labs(
    title = "Per-settlement response changes smoothly with the threshold",
    subtitle = "Dashed line marks the preferred p-lit < 0.40 not-up proxy",
    x = "Daily threshold: p-lit below gate",
    y = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

```{r settlement-slope-distribution-plot, fig.cap="Population-weighted settlement-level slope distributions for selected gates. Histograms are trimmed to each gate's weighted 1st-99th percentile range; dashed lines show the population-weighted mean slope."}
ggplot(slope_plot_data, aes(x = slope_pp, weight = population)) +
  geom_histogram(bins = 42, fill = "#DDAA33", color = "white", linewidth = 0.15, alpha = 0.86) +
  geom_vline(xintercept = 0, color = "#4B5563", linewidth = 0.4) +
  geom_vline(
    data = slope_mean_lines,
    aes(xintercept = population_weighted_mean_slope_pp),
    linetype = "dashed",
    color = "#8C5A12",
    linewidth = 0.65
  ) +
  facet_wrap(~ gate_label, scales = "free_x", ncol = 2) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "The zero-response mass shrinks as the gate becomes looser",
    subtitle = "Higher gates pick up broader dimming responses, but they also become less specific in the aggregate validation",
    x = "Settlement-level effect per +10 pp Eskom MLR",
    y = "Population mass"
  ) +
  theme_minimal(base_size = 10)
```

The settlement-level graph does change with the gate. At low gates, a large population mass has zero slopes because many settlements are almost never classified strict or mostly dark. As the threshold rises, more settlements enter the response margin and the population-weighted mean slope increases. This helps explain why the aggregate validation remains stable through `0.50` and `0.60`: load reduction is moving many otherwise bright settlements into partial-not-up states, not only into full darkness.

That does not mean the highest gates are better. The settlement-level response broadens, but the aggregate validation fit still deteriorates at very high gates. The high-gate distributions show more settlements responding, while the R2 plots above show that those responses are less specifically tied to Eskom MLR once the threshold approaches `0.90` or `0.95`.

# Does This Undermine the Metric?

No. It supports the interpretation of `p_lit < 0.40` as a practical broad not-up threshold.

The high gates remain statistically significant, but the validation fit deteriorates strongly as the gate rises. That is exactly what we should expect if high gates increasingly include ordinary dim-but-serviceable or partially lit states. In the demand-weighted simple bridge, R2 falls from `r fmt_num(preferred_demand_040$r_squared, 3)` at `0.40` to `r fmt_num(simple_results$r_squared[simple_results$weighting == "Demand" & simple_results$threshold == 0.80], 3)` at `0.80`, `r fmt_num(simple_results$r_squared[simple_results$weighting == "Demand" & simple_results$threshold == 0.90], 3)` at `0.90`, and `r fmt_num(simple_results$r_squared[simple_results$weighting == "Demand" & simple_results$threshold == 0.95], 3)` at `0.95`.

Therefore:

- `p_lit < 0.05` remains the cleanest blackout-like signal.
- `p_lit < 0.20` remains the mostly-dark sensitivity.
- `p_lit < 0.40` remains defensible as the broad not-up / degraded-service proxy for PyPSA.
- thresholds above `0.50` should not be used as the preferred ENS proxy, because they remain significant but become progressively less specific to load reduction.

The validation signal is not fragile. The sensitivity shows a sensible degradation curve: strong at low-to-moderate gates, still present but increasingly diluted at very high gates.
